A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-11-13 thesis/thesis.tex
apply more comments

\pagebreak
\section{Related Work}
In this section, we look at previous lock-free queues. In the worst-case, step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or a dequeue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

@@

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom. Their step complexity is $\Omega(p)$ because of the helping mechanism.

@@

\begin{lemma}[Amortized Time Analysis]
The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
\end{lemma}
\begin{proof}
If we split the \nf{DoublingSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
\end{proof}

@@

\paragraph{Attiya--Fouren Lower Bound}
As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $O(\log\log p)$. Our amortized complexity is $O(\log^2p+\log q)$. It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower bound.

>>>>

\pagebreak
\section{Related Work}
In this section, we look at previous lock-free queues. The worst-case step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or a dequeue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

@@

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom. Their worst-case step complexity is $\Omega(p)$ because of the helping mechanism.

@@

\begin{lemma}[Amortized Worst-case Analysis]
The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in the worst-case, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
\end{lemma}
\begin{proof}
If we split the \nf{DoublingSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
\end{proof}

\begin{corollary}
    If the number of concurrent processes in an execution is less than $c$, then each \nf{GetEnqueue} takes $O(\log c)$ steps because each block has at most $c$ operations. So the total amortized worst-case step complexity would be $O(\log p\log c + \log q)$. Furthermore, in a sequential execution where $c=1$, the step complexity of our algorithm would be $\Theta(\log p + \log q)$.
\end{corollary}

@@

\paragraph{Attiya--Fouren Lower Bound}
As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $O(\log\log p)$. Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$. It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower bound.
